Bayesian integration of networks without gold standards

### 2011-09-05 src/main/latex/paper/paper.tex
updated paper figures with synthetic GI data
paper: some more results, some more literature

Inferring confidence assessments for semantically integrated datasets is challenging, because each single type of association must be scored separately. However, gold standards only exist for some types of associations. Therefore, it is desirable to find a method that can infer confidence measures on biological networks in the absence of a gold standard, that is, based only on the existing experimental data. So far, only partial solutions to this problem exist. These solutions infer confidence measures for a specific type of association only, for example, protein-protein interactions \citep{bader_gaining_2004}.\note{Find more}

@@

The simulated true graph is created as a scale-free graph using a preferential attachment algorithm, since most molecular-biological graphs have been shown to be approximately scale-free \citep{jeong_lethality_2001, eisenberg_preferential_2003}. A description of the algorithm can be found in the supplementary material.
\note{Mention how this does not match the algorithm's definition of the prior distribution of $G$, thus making it more challenging for the algorithm to perform.}%TODO

@@

\subsection{Semantic integration}

In addition to the evaluation on artificial data it is important to analyse the method's performance on real biological data. To this end we have used the semantic data integration system Ondex \citep{koehler_graph-based_2006} to gather as much data as possible on the \species{Saccharomyces cerevisiae} protein-protein interaction (PPI) network. A new PSI-MI 2.5 XML \citep{kerrien_broadening_2007} parser was written to import data from the BioGRID \citep{breitkreutz_biogrid_2008}, MINT \citep{chatr-aryamontri_mint:_2007}, IntAct \citep{hermjakob_intact:_2004} and MIPS-MPACT \citep{guldener_mpact:_2006}. The resulting Ondex dataset represented proteins, their interactions, the experiments in which the interactions have been observed and the publications in which the experiments were described. A semantic merger method was developed to clean the dataset of its redundancies. Further details on the semantic integration procedure can be found in the supplementary material.

@@

\subsection{Application to physical and genetic interactions}

Figure \ref{ppi+gi} shows histograms of the existence probabilities for physical and genetic interactions from the semantically integrated dataset as calculated by the method presented here. A share of 39.03\% of the experimentally reported protein-protein interactions have been assigned probabilites less than 0.1. This percentage is lower than what could be expected according to the projections of Hart and colleagues \citep{hart_how_2006}.

\begin{figure}[!tpb]
\centerline{\includegraphics[width=.5\textwidth]{fig2.pdf}}
\caption{Histogram of existence probabilities for the interactions in the integrated dataset. Top: Protein-protein interactions, bottom: genetic interactions.}
\label{ppi+gi}
\end{figure}

%When given a mixed set of HTP and LTP experiments, ...\note{TODO: Implement LTP simulation}

\section{Discussion}
\note{
\begin{itemize}
  \item Excellent performance despite giving competitors head start with unbiased gold standards and accurate fudge factors, which wouldn't be available normally.
  \item Other possible performance metrics (e.g. ROC curves or weighted sums)
  \item Discuss discrepancy between Hart et al and the PPI network
  \item Future work: Evaluate on simulated low-throughput
  \item Future work: Further evaluation of the algorithm using the integrated dataset
  \begin{itemize}
    \item Measure functional annotation accuracy with leave-one-out
  \end{itemize}
\end{itemize}
}

@@

\note{TODO: Fix capitalization and italics in BibTeX file}

>>>>

Inferring confidence assessments for semantically integrated datasets is challenging, because each single type of association must be scored separately. However, gold standards only exist for some types of associations. Therefore, it is desirable to find a method that can infer confidence measures on biological networks, preferrably in the absence of a gold standard, that is, based only on the existing experimental data. So far, only partial solutions to this problem exist. These solutions infer confidence measures for a specific type of association only, for example, protein-protein interactions \citep{bader_gaining_2004,braun_experimentally_2009,venkatesan_empirical_2009}.

@@

The simulated true graph is created as a scale-free graph using a preferential attachment algorithm, since most molecular-biological graphs have been shown to be approximately scale-free \citep{jeong_lethality_2001, eisenberg_preferential_2003}. A description of the algorithm can be found in the supplementary material.
Not only does the choice of scale free background graphs more closely match the topology of real biological graphs; it also poses an additional challenge for the new algorithm, since such graphs break with the assumption of a uniform prior distribution for $G$.

@@

\subsection{Application to biological networks}

In addition to the evaluation on artificial data we applied the new method to semantically integrated biological data. We used the data integration system Ondex \citep{koehler_graph-based_2006} to gather as much data as possible on the \species{Saccharomyces cerevisiae} protein-protein interaction (PPI) and genetic interaction (GI) network. W have imported all\species{S. cerevisiae} data from the BioGRID \citep{breitkreutz_biogrid_2008}, MINT \citep{chatr-aryamontri_mint:_2007}, IntAct \citep{hermjakob_intact:_2004} and MIPS-MPACT \citep{guldener_mpact:_2006} via the PSI-MI 2.5 XML \citep{kerrien_broadening_2007} format. The resulting Ondex dataset represented proteins and genetic features, as well as their interactions, the experiments in which the interactions have been observed and the publications in which the experiments were described. To identify and interlink equivalent entries we implemented a semantic merger method. Finally, we summarized small low-throughput experiments (<20 interactions) into larger groups according to their experimental type, since low sample sizes can be expected to lead to very unprecise error rate estimates. Further details on the semantic integration procedure can be found in the supplementary material.

Having established a semantic knowledge network of physical and genetic interactions within \species{S. cerevisiae} We executed the new algorithm over all 51,006 contained physical interactions (PSI-MI ontology ID MI:0915) and all 15,006 contained synthetic genetic interactions (PSI-MI ontology ID MI:0794), as an exemplary type of genetic interaction.

@@

\subsection{Interaction probabilities for \species{S. cerevisiae}}

Figure \ref{ppi+gi} shows histograms of the existence probabilities for physical interactions and synthetic genetic interactions from the semantically integrated dataset as calculated by the method presented here. Regarding physical interactions, 20,180 (39.03\%) of the experimentally reported interactions have been assigned probabilites less than 0.1. A total of 22090 (43.3\%) have been assigned probabilites greater than 0.9. The remaining interactions are widely distributed around a small peak near probability 0.3.

\note{TODO: Output error rates and compare with values from \citep{dhaeseleer_estimating_2004} and \citep{hart_how_2006}}

When looking at the probability distribution over synthetic genetic interactions it is apparent that with 14,386 (95.8\%) a vast majority of interactions have been assigned values of less than 0.5. \note{Potential explanation: Extremely poor coverage. TODO: run coverage analysis in Ondex to confirm.} Of these, 8086 (53.9\%) have a probability less than 0.1. Only 230 (1.5\%) have a probability greater than 0.9.

%\citep{hart_how_2006} summarizes literature about PPI interactome size projections as between 37,800-75,500.

\begin{figure}[!tpb]
\centerline{\includegraphics[width=.5\textwidth]{fig2.pdf}}
\caption{Histogram of existence probabilities for the interactions in the integrated dataset. Top: Protein-protein interactions, bottom: synthetic genetic interactions.}
\label{ppi+gi}
\end{figure}

\section{Discussion}
\note{
\begin{itemize}
  \item Excellent performance despite giving competitors head start with unbiased gold standards and accurate fudge factors, which wouldn't be available normally.
  \item Other possible performance metrics (e.g. ROC curves or weighted sums)
  \item Future work: Evaluate on simulated low-throughput
  \item Future work: Further evaluation of the algorithm using the integrated dataset
  \begin{itemize}
    \item Measure functional annotation accuracy with leave-one-out
  \end{itemize}
  \item Future work: include confidence values given by original experiments. (Discuss problem of missing standards)
\end{itemize}
}

@@

\note{TODO: Fix capitalization and italics in BibTeX file (Adding items from Zotero will undo this, so wait until the end)}
